Compositional Equivalences Based on Open pNets

### 2022-10-05 2016-Weak-FH-Bisim/WeakBisim.tex
comments up to conclusion

\subsubsection*{Open and compositional systems}
%\TODO{retravailler le paragraphe ci dessous: faire ref a Facs-2013 (reconfig ds pNets) - virer 14 (probabilistic) ?} \cite{GasparHM:2013}
In
\cite{Johnson:2013:CBSE,Johnson:2014:qosa}, the authors investigate several methodologies for the compositional verification of software systems, with the aim to verify reconfigurable component systems. To improve scaling and compositionality, the authors  decompose the verification problem that is to be resolved by a SMT (satisfiability modulo theory) solver into independent sub-problems on independent sets of variables. These works clearly highlight the interest of incremental and compositional verification in a very general setting. In our own work on open pNets, adding more structure to the composition model, we show how to enforce a compositional proof system that is more versatile than independent sets of variables as the composition is  structured and allows arbitrary synchronisations between sub-entities. Our theory has also been encoded into an SMT solver and it would be interesting to investigate how the examples of evolving systems studied by Johnson et al. could be encoded into pNet and verified by our framework. However, the models of Johnson et al. are quite different from ours, in particular they are much less structured, and translating them is clearly outside the scope of this article.
In previous work~\cite{GasparHM:2013}, we also have shown how (closed) pNet models could be used to encode and verify finite instances of reconfigurable component systems. 
%Additionally, the more compositional constraint resolution provided by the authors could also be interesting to make our SMT proofs scale better (inside a single pNet for example), the integration of the Z3 framework described in \cite{Johnson:2014:qosa} with our proof theory for open pNets would also be worth investigating from a practical point of view. 
%One major different however is that pNets in their current state are not much adapted to probabilistic verification contrarily to \cite{Johnson:2014:qosa} but they are better suited to verify complex behavioural properties.

@@

Methodologies for reasoning about abstract semantics of open systems can be found in \cite{BaldanBB:2002, BaldanBB:07,Dubut:20}, authors introduce  behavioural equivalences for open systems from various symbolic approaches. Working in the setting of process calculi, some close relations  exist with the work  of the authors of \cite{BaldanBB:2002,BaldanBB:07}, where both approaches are based
on some kinds of labelled transition systems. The distinguishing feature of their approach is that the transitions systems are labelled with logical formulae that provides  an abstract characterization of the structure that a hole must possess and of the actions it can perform in order to allow a transition to fire. Logical formulae are suitable formats that capture the general class of components that can act as the placeholders of the system during its evolution. In our approach we purposely leave the algebra of action terms undefined but the only operation we allow on action of holes is the comparison with other actions. Defining properly the interaction between a logical formulae in the action and the logics of the pNet composition seems very difficult.

Among the approaches for modelling open systems, one can cite  \cite{Beohar:2020} that uses transition conditions depending on an external environment, and introduce bisimulation relations based on this approach. The approach of  \cite{Beohar:2020} is 
  highly based on logics and their bisimulation theory richer in this aspect, while our theory is highly structural and focuses on relation between structure and equivalence. Also, we see composition as a structural operation putting systems together, and do not focus on the modelisation of an unknown outside world. Overall we believe that the two approaches are complementary but checking the compatibility of the two different bisimulation theories is not  trivial.

@@

There is also a clear relation with the seminal works on rule formats for Structured Operational Semantics, e.g. DeSimone format, GSOS, and conditional rules with or without negative premisses \cite{deSimone85,Bloom:1995,GROOTE1992202,VANGLABBEEK2004229}.  The Open pNets model provides a way to define operators similar to these rules formats, but with quite different aim and approach. A formal comparison would be interesting, though not trivial. What we can say easily is that: the pNet format syntactically encompasses both DeSimone, GSOS, and conditional premisses rules. Then our compositionality result is more powerful than their classical results, but this is not a surprise, as we rely on a (sufficient) syntactic hypothesis on a particular system, rather than the general rules defining an operator.  Last, we intentionally do not accept negative premisses, that would be more  to put into practice in our implementation. an extension could be studied in future work.

\subsubsection*{Symbolic and data-sensitive systems}
 As mentioned in the \emph{Introduction}, the work that brought us a lot of inspirations are those of Lin et al. \cite{IngolfsdottirL:2001,HennessyLin:TCS95,Lin:96}. They developed the theory of symbolic transition graphs (STG), and the associated symbolic (early and late, strong and weak) bisimulations, they also study STGs with assignments as a model for message-passing processes. Our work extends these in several ways: first our models are compositional, and our bisimulations come with effective conditions for being preserved by pNet composition (i.e. congruent), even for the weak version. This result is more general than the bisimulation congruences for value-passing CCS in \cite{IngolfsdottirL:2001}. Then our settings for management of data types are much less restrictive, thanks to our use of satisfiability engines, while Lin's algorithms were limited to data-independent systems.

@@

Research related to the keyword "Symbolic Bisimulation" refer to two very different domains, namely BDD-like techniques for modelling and computing finite-state bisimulations, that are not related to our topic; and symbolic semantics for data-dependant or high-order systems, that are very close in spirit to our approach. In this last area, we can mention
Calder's work \cite{calder2001symbolic}, that  defines a symbolic semantic for full Lotos, with a symbolic bisimulation over it;
Borgstrom et al., Liu et al, Delaune et al. and Buscemi et al. providing symbolic semantic and equivalence for different variants of pi calculus respectively \cite{BorgstromBN04,DelauneKR:2007,LiuL:2010,BuscemiM:2008}; and more recently
 Feng et al. provide a symbolic bisimulation for quantum processes \cite{feng2014symbolic}.
All the above works, did not give a complete approach for verification, and the models on which these works based are definitely different from ours.

@@

\section{Conclusion and discussion}
\label{section:conclusion}
 pNets (Parameterised Networks of Automata) is a formalism adapted to the representation of the behaviour of a parallel or distributed systems. One  strength of pNets is their parameterised nature, making them adapted to the representation of systems of arbitrary size, and making the modelling of parameterised system possible. Parameters are also crucial to reason on interaction protocols that can address one entity inside an indexed set of processes. pNets have been successfully used to represent behavioural specification of parallel and distributed components and verify their correctness~\cite{AmeurBoulifa2017,HenrioKSM:2016}. VCE is the specification and verification platform that uses pNets as an intermediate representation.

Open pNets are pNets with holes; they are adapted to represent processes parameterised by the behaviour of other processes, like composition operators or interaction 
protocols that synchronise the actions of processes that can be plugged afterwards. Open
pNets are hierarchical composition of automata with holes and parameters. We
defined here a semantics for open pNets and a complete bisimulation theory for them. 
The semantics of open pNets relies on the definition of open automata that are automata with holes and parameters, but no hierarchy. Open automata are somehow labelled transition systems with parameters and holes, a notion that is useful to define semantics, but makes less sense when modelling a system, compared to pNets. To be precise, it is on open automata that we define our FH-bisimilarity relations.

This article defines a strong and a weak bisimulation relation that are adapted to parameterised systems and hierarchical composition. Our FH-bisimulation principle handles pNet parameters in the sense that two states might be or not in relation depending on the value of parameters. Our strong  FH-bisimilarity is compositional by nature in the sense that FH-bisimilarity is maintained when composing processes. We also identified a simple and realistic condition on the semantics of non-observable actions that allows weak FH-bisimilarity to be also compositional. Overall we believe that this article paved the way for a solid theoretical foundation for compositional verification of parallel and distributed systems.

>>>>

\subsubsection*{Open and compositional systems}
%\TODO{retravailler le paragraphe ci dessous: faire ref a Facs-2013 (reconfig ds pNets) - virer 14 (probabilistic) ?} \cite{GasparHM:2013}
In
\cite{Johnson:2013:CBSE,Johnson:2014:qosa}, the authors investigate several methodologies for the compositional verification of software systems, with the aim to verify reconfigurable component systems. To improve scaling and compositionality, the authors  decompose the verification problem that is to be resolved by a SMT (satisfiability modulo theory) solver into independent sub-problems on independent sets of variables. These works clearly highlight the interest of incremental and compositional verification in a very general setting. In our own work on open pNets, adding more structure to the composition model, we show how to enforce a compositional proof system that is more versatile than independent sets of variables as the composition is  structured and allows arbitrary synchronisations between sub-entities. Our theory has also been encoded into an SMT solver and it would be interesting to investigate how the examples of evolving systems studied by Johnson et al. could be encoded into pNet and verified by our framework. However, the models of Johnson et al. are quite different from ours, in particular they are much less structured, and translating them is clearly outside the scope of this article.

In previous work~\cite{GasparHM:2013}, we also have shown how (closed) pNet models could be used to encode and verify finite instances of reconfigurable component systems. 
%Additionally, the more compositional constraint resolution provided by the authors could also be interesting to make our SMT proofs scale better (inside a single pNet for example), the integration of the Z3 framework described in \cite{Johnson:2014:qosa} with our proof theory for open pNets would also be worth investigating from a practical point of view. 
%One major different however is that pNets in their current state are not much adapted to probabilistic verification contrarily to \cite{Johnson:2014:qosa} but they are better suited to verify complex behavioural properties.

@@

Methodologies for reasoning about abstract semantics of open systems can be found in \cite{BaldanBB:2002, BaldanBB:07,Dubut:20}, authors introduce  behavioural equivalences for open systems from various symbolic approaches. Working in the setting of process calculi, some close relations  exist with the work  of the authors of \cite{BaldanBB:2002,BaldanBB:07}, where both approaches are based
on some kinds of labelled transition systems. The distinguishing feature of their approach is that the transitions systems are labelled with logical formulae that provides  an abstract characterization of the structure that a hole must possess and of the actions it can perform in order to allow a transition to fire. Logical formulae are suitable formalisms that capture the general class of components that can act as the placeholders of the system during its evolution. In our approach we purposely leave the algebra of action terms undefined but the only operation we allow on action of holes is the comparison with other actions. Defining properly the interaction between a logical formulae in the action and the logics of the pNet composition seems very difficult.
\TODO{Rabea: look for "quotienting techniques"}

Among the approaches for modelling open systems, one can cite  \cite{Beohar:2020} that uses transition conditions depending on an external environment, and introduce bisimulation relations based on this approach. The approach of  \cite{Beohar:2020} is 
  highly based on logics and their bisimulation theory is richer than ours in this aspect, while our theory is highly structural and focuses on relation between structure and equivalence. Also, we see composition as a structural operation putting systems together, and do not focus on the modelling of an unknown outside world. Overall we believe that the two approaches are complementary but comparing precisely the two different bisimulation theories is not  trivial.

@@

There is also a clear relation with the seminal works on rule formats for Structured Operational Semantics, e.g. DeSimone format, GSOS, and conditional rules with or without negative premisses \cite{deSimone85,Bloom:1995,GROOTE1992202,VANGLABBEEK2004229}.  The Open pNets model provides a way to define operators similar to these rules formats, but with quite different aim and approach. A formal comparison would be interesting, though not trivial. What we can say easily is that: the pNet format syntactically encompasses  De Simone, GSOS, and conditional premisses rules. Then our compositionality result is more powerful than their classical results, but this is not a surprise, as we rely on a (sufficient) syntactic hypothesis on a particular system, rather than the general rules defining an operator.  Last, we intentionally do not accept negative premisses, that would be more  to put into practice in our implementation. This extension could be studied in future work.

\subsubsection*{Symbolic and data-sensitive systems}
 As mentioned in the \emph{Introduction}, we were substantially inspired by the works  of Lin et al. \cite{IngolfsdottirL:2001,HennessyLin:TCS95,Lin:96}. They developed the theory of symbolic transition graphs (STG), and the associated symbolic (early and late, strong and weak) bisimulations. Moreover, they studied STGs with assignments as a model for message-passing processes. Our work extends those contributions in several ways: first our models are compositional, and our bisimulations come with effective conditions for being preserved by pNet composition (i.e. congruent), even for the weak version. This result is more general than the bisimulation congruences for value-passing CCS in \cite{IngolfsdottirL:2001}. Then our settings for management of data types are much less restrictive, thanks to our use of satisfiability engines, while Lin's algorithms were limited to data-independent systems.

@@

Research related to the keyword "Symbolic Bisimulation" refer to two very different domains, namely BDD-like techniques for modelling and computing finite-state bisimulations, that are not related to our topic; and symbolic semantics for data-dependant or high-order systems, that are very close in spirit to our approach. In this last area, we can mention
Calder's work \cite{calder2001symbolic}, that  defines a symbolic semantic for full Lotos, with a symbolic bisimulation over it;
Borgstrom et al., Liu et al, Delaune et al. and Buscemi et al. providing symbolic semantics and equivalence for different variants of pi calculus respectively \cite{BorgstromBN04,DelauneKR:2007,LiuL:2010,BuscemiM:2008}; and more recently
 Feng et al. provide a symbolic bisimulation for quantum processes \cite{feng2014symbolic}.
\TODO{check next sentence}
All the above works are based on models  definitely different from ours, and none of them allows system to be as much parameterised as open pNets; this additional expressiveness due to the open and symbolic nature of our constructs.

@@

\section{Conclusion and discussion}
\label{section:conclusion}
 pNets (Parameterised Networks of Automata) is a formalism adapted to the representation of the behaviour of  parallel or distributed systems. One  strength of pNets is their parameterised nature, making them suitable for to the representation of systems of arbitrary size, and making the modelling of parameterised system possible. Parameters are also crucial to reason on interaction protocols that can address one entity inside an indexed set of processes. pNets have been successfully used to represent behavioural specification of parallel and distributed components and verify their correctness~\cite{AmeurBoulifa2017,HenrioKSM:2016}. VCE is the specification and verification platform that uses pNets as an intermediate representation.
\TODO{Eric: short mentions and links to our tools and examples papers}

Open pNets are pNets with holes; they are adapted to represent processes parameterised by the behaviour of other processes, like composition operators or interaction 
protocols that synchronise the actions of processes that can be plugged afterwards. Open
pNets are hierarchical composition of automata with holes and parameters. We
defined here a semantics for open pNets and a complete bisimulation theory for them. 
The semantics of open pNets relies on the definition of open automata that are automata with holes and parameters, but no hierarchy \TODO{Ludo: flatten}. Open automata are somehow labelled transition systems with parameters and holes, a notion that is useful to define semantics, but makes less sense when modelling a system, compared to pNets. To be precise, it is on open automata that we define our FH-bisimilarity relations.

This article defines a strong and a weak bisimulation relation that are adapted to parameterised systems and hierarchical composition. Our FH-bisimulation principle handles pNet parameters in the sense that two states might be or not in relation depending on the value of parameters. Our strong  FH-bisimilarity is compositional by nature in the sense that FH-bisimilarity is maintained when composing processes. We also identified a simple and realistic condition on the semantics of non-observable actions that allows weak FH-bisimilarity to be also compositional \TODO{Eric: recall on rule formats}. Overall we believe that this article paved the way for a solid theoretical foundation for compositional verification of parallel and distributed systems.
